---
title: "Bias-estimation demo"
author: "Michael McLaren"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
    highlight: tango
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(tidyverse)
library(ggbeeswarm)
devtools::load_all(here::here())
```

```{r}
set.seed(1)
```

Bias for 5 fake taxa:
```{r}
K <- 5
taxa <- paste0("T", seq(K))
bias <- tibble(Taxon = taxa, Bias = exp(rnorm(K, 0, 2))) %>%
    mutate(
        Bias = center_elts(Bias),
    )
bias
```

## Experiment where all samples contain all taxa

Let's first consider an experiment where all samples have all taxa. I will draw
the actual abundances as iid from an exponential distribution, and take the
residual error to be iid log-normal with an sd on the log scale of 0.5.
```{r}
N <- 10
samples <- paste0("S", seq(N))
exp1 <- crossing(Sample = samples, Taxon = taxa) %>%
    left_join(bias, by = "Taxon")
exp1 <- exp1 %>%
    mutate(
        Actual = rexp(n(), 1),
        Noise = exp(rnorm(n(), sd = 0.5)),
        Observed = Actual * Bias * Noise
    ) %>%
    group_by(Sample) %>%
    mutate_at(vars(Actual, Noise, Observed), center_elts) %>%
    ungroup
```
The `center_elts` function normalizes the vectors by the geometric mean of
their elements.

These samples have varying actual compositions, but the compositional
differences (Observed/Actual) will appear consistent when normalized in this
manner:
```{r}
p <- ggplot(exp1, aes(Taxon, Observed / Actual)) + 
    geom_point(aes(y=Bias), shape = 3, size = 5, color = "darkgrey") +
    geom_quasirandom() +
    scale_y_log10()
p
```

The systematic deviation from 1 corresponds to the Bias, with the residual
variation due to the Noise factors.


The `estimate_bias` function requires a data frame with Sample, Actual, and
Observed columns. It does not depend on how the abundances are normalized (e.g.
as proportions, or as above are both fine). It has two methods, the "center"
method, which estimates the bias by the compositional mean of the observed
errors, and the "rss" method, which estimates the bias by minimizing the
sum of squares of the residual errors as measured in the Aitchison geometry.
Here, both methods can be applied and give the same answer
```{r}
est1 <- bias %>%
    left_join(estimate_bias(exp1, method = "center"), by = "Taxon") %>%
    left_join(estimate_bias(exp1, method = "rss"), by = "Taxon")
est1
```
and we can see that the estimate is close to the true Bias,
```{r}
p + geom_point(data = est1, aes(y=Bias_est.x), shape = 3, size = 5, 
        color = "darkred")
```

This estimate is also the MLE under Aitchison's logistic multivariate-normal
error model, which corresponds to assuming the observed error after log-ratio
transform to be multivariate normal. This is clear from the fact that we are
simply computing the sample mean in log-ratio space, and we can check this
directly by getting the MLE with the `mvnmle` package:
```{r, warning=FALSE}
tb <- exp1 %>%
    mutate(Observed_error = Observed / Actual) %>%
    select(Sample, Taxon, Observed_error) %>%
    mutate_by(Sample, Observed_error = clr(Observed_error)) %>%
    spread(Taxon, Observed_error) %>%
    select(-Sample)
fit <- mvnmle::mlest(tb, iterlim = 1000)
est1 <- est1 %>%
    mutate(Bias_est.z = fit$muhat %>% exp %>% center_elts)
est1
```

## Experiment where samples contain different taxa

Next, consider a case where different samples have different taxa, by 
removing ~2/3 of taxa randomly from the first experiment and recomputing the
observed compositions,
```{r}
exp2 <- exp1 %>%
    mutate(
        Actual = Actual * rbernoulli(n(), p = 0.66),
        Observed = Actual * Bias * Noise
    )
exp2
```
Note, we no longer center the abundance vectors, since the geometric means for
samples without all taxa will be zero. 

To visualize bias in a composition-independent manner, we instead can compute
pairwise ratios:
```{r}
ratios <- crossing(Sample = samples, Taxon.x = taxa, Taxon.y = taxa) %>%
    filter( Taxon.x < Taxon.y) %>%
    left_join(exp2, by = c("Sample", "Taxon.x" = "Taxon")) %>%
    left_join(exp2, by = c("Sample", "Taxon.y" = "Taxon")) %>%
    mutate(Pair = paste(Taxon.x, Taxon.y, sep = ":"))
ratios <- ratios %>%
    filter(Actual.x, Actual.y) %>%
    mutate(
        Observed = Observed.x / Observed.y,
        Actual = Actual.x / Actual.y,
        Bias = Bias.x / Bias.y
    )
```
```{r}
p <- ggplot(ratios, aes(Pair, Observed / Actual)) +
    geom_point(aes(y=Bias), shape = 3, size = 5, color = "darkgrey") +
    geom_quasirandom() +
    scale_y_log10()
p
```

The "center" method of `estimate_bias` gives NaN's because the compositional
differences between the Observed and Actual abundance vectors contain elements
with 0/0. But the "rss" method still gives a sensible answer.
```{r}
est2 <- bias %>%
    left_join(estimate_bias(exp2, method = "center"), by = "Taxon") %>%
    left_join(estimate_bias(exp2, method = "rss"), by = "Taxon")
est2
```

We can check the fit by looking at the predicted error in the pairwise ratios,
```{r}
est2.1 <- est2 %>%
    select(Taxon, Bias_est = Bias_est.y)
ratios <- ratios %>%
    left_join(est2.1, by = c("Taxon.x" = "Taxon")) %>%
    left_join(est2.1, by = c("Taxon.y" = "Taxon")) %>%
    mutate(Bias_est = Bias_est.x / Bias_est.y)
```
```{r}
p +
    geom_point(data = ratios, aes(y=Bias_est), shape = 3, size = 5, 
        color = "darkred")
```

The `mvnmle` package can still be used to perform maximum likelihood estimation
in this situation. Rather than computing the `clr`, we take the log of each
taxon as if taking the `alr` to an imaginary reference taxon assumed to have
an observed error (Observed/Actual) equal to 1.
```{r}
tb <- exp2 %>%
    mutate(Observed_error = Observed / Actual) %>%
    select(Sample, Taxon, Observed_error) %>%
    mutate_by(Sample, Observed_error = log(Observed_error)) %>%
    spread(Taxon, Observed_error) %>%
    select(-Sample)
tb
```
Under my understanding, the NaN's will be treated as missing at random (MAR)
and so not influence the likelihood (which is what we want).
```{r, warning=FALSE}
fit <- mvnmle::mlest(tb, iterlim = 1000)
est2 <- est2 %>%
    mutate(Bias_est.z = fit$muhat %>% exp %>% center_elts)
est2
```
This estimate differs somewhat from the "rss" result. It is also quite unstable
for `iterlim` lower than ~1000 in this small dataset, and seems to become much
more unstable in slightly larger datasets or datasets with more missing taxa,
so I'm not sure if I trust the difference seen here.

## Experiment where samples have disjoint sets of taxa

Now suppose the first 5 samples have the first 3 taxa, and the last 5 samples
have the last 2 taxa:
```{r}
exp3 <- exp1 %>%
    mutate(
        Actual = case_when(
            Sample <= "S5" & Taxon <= "T3" ~ Actual,
            Sample <= "S5" & Taxon > "T3" ~ 0,
            Sample > "S5" & Taxon <= "T3" ~ 0,
            Sample > "S5" & Taxon > "T3" ~ Actual,
        ),
        Observed = Actual * Bias * Noise
    )
exp3
```
In this case, the 5 taxa form two sets, such that we can estimate the bias
within sets but not between them, since there are no samples with taxa from
both sets. The `estimate_bias` function's "rss" method will still work, but we
can't trust the relative efficiencies between sets of taxa, only within sets:
```{r}
est3 <- bias %>%
    left_join(estimate_bias(exp3, method = "rss"), by = "Taxon") %>%
    mutate(Set = case_when(
            Taxon <= "T3" ~ "Set1",
            Taxon > "T3" ~ "Set2",
        ))
est3 <- est3 %>%
    mutate_by(Set,
        Bias.in_set = center_elts(Bias),
        Bias_est.in_set = center_elts(Bias_est)
    )
est3
```

