---
title: "Costea2017 analysis"
author: Michael McLaren
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
    highlight: tango
---

# R setup 

```{r setup, include=FALSE}
knitr::opts_knit$set(progress = TRUE, verbose = TRUE)
# Global chunk options
knitr::opts_chunk$set(
    cache = TRUE, autodep = TRUE,
    include = TRUE, echo = TRUE,
    warning = TRUE, message = FALSE, 
    fig.width = 6, fig.height = 4
)
```

Run with `SAVE_FIGURES = TRUE` to save figures in `figures/`. 
```{r}
SAVE_FIGURES = TRUE
fig_path <- file.path(here::here(), "figures")
```

## Libraries and paths

```{r load_packages}
library(tidyverse)
library(ggthemes)
library(cowplot)
library(ggbeeswarm)
# This package
devtools::load_all(here::here())
```

## Plot setup

Options for `ggplot`:
```{r}
base_theme <- theme_tufte() + 
    theme(
        text = element_text(size=9, family = ""),
        legend.position = "none"
        )
base_theme0 <- theme_grey() + 
    theme(
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank()
        )
tax_theme <- theme(
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1), 
        axis.title.x = element_blank())

update_geom_defaults("point", list(size = 1))
update_geom_defaults("text", list(size = 2.5))
update_geom_defaults("hline", list(color = "grey"))
```

```{r}
costea.shape_scale <- scale_shape_manual(
    breaks = c("I", "M"), 
    values = c(16, 1),
    labels = c("Fecal", "Mock only"))

# colors.protocol <- scales::hue_pal()(3)
# colors.protocol <- c("H" = "#F8766D", "Q" = "#00BA38", "W" = "#619CFF")
# names(colors.protocol) <- c("H", "Q", "W")
# Or, to match the Costea paper:
# names(colors.protocol) <- c("W", "H", "Q")
# colors.protocol <- c(colors.protocol, Actual = "black")

# Color-blind friendly colors from ColorBrewer2
colors.protocol <- c(
    H = "#1b9e77", 
    W = "#d95f02", 
    Q = "#7570b3", 
    Actual = "black")
```

# Data setup

Sample metadata:
```{r}
data("costea2017_sample_data")
sam <- costea2017_sample_data
sam
```
Expected mock community composition, obtained from the supplementary data (see
`data-raw/costea2017.R`)
```{r}
data("costea2017_mock_composition")
mock <- costea2017_mock_composition
```
The FACS measurements are the "bacterial cells" columns; the measurements
without endospores were used as the actual composition, and replicate
measurements were averaged (personal communication with Paul Costea); 
```{r}
mock <- mock %>%
    select(Taxon, FACS = "bacterial cells in spike in Mix") %>%
    group_by(Taxon) %>%
    summarize_at(vars(FACS), mean)
mock <- mock %>%
    arrange(desc(FACS))
mock
```
Note the difference in order for three species from Costea2017 Figure 6. None
of the available columns give the ordering seen in their figure, indicating a
mislabeling in the figure or in the provided spreadsheet. However, the
mislabeling does not affect our main conclusions; we can analyze the data as if
the spreadsheet is correct.

Metaphlan2 profiles:
```{r}
data("costea2017_metaphlan2_profiles")
profiles <- costea2017_metaphlan2_profiles
profiles %>% corner
```
The samples are named here by their ENA run accession; let's rename to the
new sample names.
```{r}
new_names <- c("Clade", 
    sam$Sample[match(colnames(profiles)[-1], sam$Run_accession)])
colnames(profiles) <- new_names
```

Two samples, QA and QB, are two extra fecal samples that were only sequenced by
protocol Q and so we won't be using for our analysis. We also drop unneeded
sample variables.
```{r}
sam <- sam %>%
    filter(!(Sample %in% c("QA", "QB"))) %>%
    select(Sample, Protocol, Individual)
profiles <- profiles[, c("Clade", sam$Sample)]
```

Taxonomy from the metaphlan2 clade names:
```{r}
tax_ranks <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus",
    "Species", "Strain")
rank_letters <- c("k", "p", "c", "o", "f", "g", "s", "t")
tax_pattern  <- paste0("(?:", rank_letters, "__(\\w+))?") %>%
    paste(collapse = "\\|?")
tax <- profiles %>%
    select(Clade) %>%
    extract(Clade, tax_ranks, regex = tax_pattern, remove = FALSE)
tax
```

Species table:
```{r}
st <- profiles %>%
    left_join(tax, by = "Clade") %>%
    filter(!is.na(Species), is.na(Strain)) %>%
    select(-Strain)
st
```
Species names have the form `Genus_species` and are unique,
```{r}
st$Species %>% head
st$Species %>% anyDuplicated
```
The `Genus_unclassified` catches cases where more reads were mapped to the
genus's markers than could be accounted for by species within the genus.
We will use these species names as taxon names,
```{r}
st <- st %>%
    mutate(Clade = Species) %>%
    rename(Taxon = Clade)
```
and join the metadata and tidy it into a tall format (one sample-taxon
observation per row)
```{r}
st <- st %>%
    gather("Sample", "Abundance", sam$Sample) %>%
    left_join(sam, by = "Sample") %>%
    select(Taxon, Sample, Abundance, colnames(sam), everything())
st
```

Metaphlan outputs estimates of the proportions * 100. We will renormalize to
proportions summing to 1 instead.
```{r}
st <- st %>%
    mutate_by(Sample, Abundance = close_elts(Abundance))
st %>%
    group_by(Sample) %>%
    summarize(sum(Abundance)) %>%
    {summary(.[,2])}
```

Output from the various kingdoms:
```{r}
st %>%
    group_by(Kingdom, Sample) %>%
    summarize(Abundance = sum(Abundance)) %>%
    spread(Kingdom, Abundance) %>%
    print(n=Inf)
```
There is very little Archaea and Eukaryote abundance, and the vast majority of
the viral abundance comes from a Salmonella phage that may be integrated into
the genome of the _Salmonella enterica_ strain in the mock community.
```{r}
st %>%
    filter(Kingdom == "Viruses") %>%
    group_by(Taxon) %>% 
    summarize(Abundance = mean(Abundance)) %>%
    arrange(desc(Abundance))
```
We will therefore restrict our analysis of sample composition to just Bacteria,
```{r}
st0 <- st %>%
    filter(Kingdom == "Bacteria") %>%
    mutate_by(Sample, Abundance = close_elts(Abundance))
```

# Composition of the samples

## Identifying contaminants

Costea et al report seeing E coli as a putative contaminant of the spike-in. We
can check by seeing what we find in the mock-only samples that shouldn't be
there:
```{r}
st0 %>%
    filter(
        Individual == "M", Abundance > 0, 
        !(Taxon %in% mock$Taxon)
        ) %>%
    group_by(Taxon) %>%
    summarize(
        Mean = mean(Abundance),
        Max = max(Abundance),
        ) %>%
    arrange(desc(Mean)) %>%
    print(n = 10)
```
It looks like the contaminant reads are mostly captured by `Shigella_flexneri`
and `Escherichia_unclassified`, with possibly some as `Escherichia_coli`. 
We will use these identifiers as putative contaminants.
```{r}
contaminants <- c("Shigella_flexneri", "Escherichia_unclassified",
    "Escherichia_coli")
```
This may lump native reads from some native Escherichia into the contaminants,
but that isn't important for our analysis.

## Taxa by source (Mock, Contaminant, Native)

Classify the taxa by source type:
```{r}
st0 <- st0 %>%
    mutate(Source = case_when(
            Taxon %in% mock$Taxon ~ "Mock", 
            Taxon %in% contaminants ~ "Contaminant",
            TRUE ~ "Native"
            )
        )
```
View the proportions of native, mock, and contaminant for each sample:
```{r}
tb.source <- st0 %>%
    group_by(Sample, Protocol, Individual, Source) %>%
    summarize(Abundance = sum(Abundance)) %>%
    ungroup()
tb.source %>%
    spread(Source, Abundance) %>%
    print(n=Inf)
```
Labeled contaminants are always a small fraction of the total sample. 
They also appear in a consistent ratio with the mock for a given protocol
except in samples Q2 and Q3,
```{r, fig.width = 4.5, fig.height = 3.5}
tb.source %>%
    spread(Source, Abundance) %>%
    ggplot(aes(Sample, Contaminant / Mock, color = Protocol, 
            label = Individual)) +
    scale_color_manual(values = colors.protocol) + 
    geom_text() +
    coord_flip()
```

And in these samples, the spike-in (including contaminants) makes up an
unusually low fraction of the total (see table above), so it is perhaps not
surprising for the ratio to be different. These facts suggest that the
`contaminants` are largely being correctly identified.

Composition by source type:
```{r}
p.source <- ggplot(tb.source,
    aes(y = Abundance, 
        x = Individual,
        # x = factor(Individual, lvls.mock_prop), 
        fill = Source)) +
    geom_bar(stat = "identity") +
    scale_y_continuous(limits=c(0,1), breaks = c(0, 0.5, 1), 
        labels = c("0", "0.5", "1")) +
    geom_rangeframe(data = tibble(y = c(0, 1), Protocol = "H"), 
        aes(y = y), sides = "l", inherit.aes = FALSE) +
    scale_fill_brewer(type = "qual") +
    facet_grid(. ~ Protocol) +
    labs(x = "Specimen", fill = "Taxon type", 
        title = "Composition of different bacterial types") +
    base_theme +
    theme(legend.position = "right")
p.source 
```

## Native taxa composition

Get the native taxa agglomerated to the Family level,
```{r}
nat <- st0 %>%
    filter(Source == "Native") %>%
    group_by_at(vars(Sample, Protocol, Individual, Kingdom:Family)) %>%
    summarize(Abundance = sum(Abundance)) %>%
    ungroup
```

```{r, eval = FALSE, include = FALSE}
ggplot(nat, aes(x = Individual, y = Abundance, fill = Family)) +
    geom_bar(stat = "identity") +
    scale_y_continuous(limits=c(0,1), breaks = c(0, 0.5, 1), 
        labels = c("0", "0.5", "1")) +
    geom_rangeframe(data = tibble(y = c(0, 1), Protocol = "H"), 
        aes(y = y), sides = "l", inherit.aes = FALSE) +
    facet_grid(. ~ Protocol) +
    labs(x = "Specimen",
        title = paste0("Composition of native bacterial taxa")
        ) +
    base_theme +
    theme(legend.position = "right") +
    labs(y = "Proportion")
```

Let's restrict to Families within a minimal abundance to simplify the figure.
First, lets check the prevalence of various families,
```{r}
prev <- nat %>%
    filter(Individual != "M") %>%
    mutate_by(Sample, Proportion = close_elts(Abundance)) %>%
    group_by(Family) %>%
    summarize(Prev = sum(Proportion > 0), 
        Min = min(Proportion), 
        Median = median(Proportion), 
        Mean = mean(Proportion), 
        Max = max(Proportion))
prev %>%
    arrange(desc(Max)) %>%
    print(n=30)
```
Let's use the families with a proportion of at least 2%.
```{r}
families <- prev %>%
    filter(Max > 0.02) %>%
    {.$Family}
nat0 <- nat %>%
    filter(Family %in% families) %>%
    arrange(Phylum, Family)
```

Create a color scale for the families that also uses hue to distinguish Phyla.
```{r}
# Color tibble
ctb <- nat0 %>%
    select(Phylum, Family) %>%
    distinct %>%
    arrange(Phylum, Family) %>%
    group_by(Phylum) %>%
    nest(Family) %>%
    rowwise() %>%
    mutate(Families = list(unlist(data)),
        Num = length(Families))
ctb <- ctb %>%
    arrange(desc(Num)) %>%
    add_column(Palette = c("Blues", "Greens", "Reds", "Oranges"))
ctb <- ctb %>%
    rowwise() %>%
    mutate(Colors = list(RColorBrewer::brewer.pal(Num, name = Palette)))

last_n <- function(x, n) {
    len <- length(x)
    x[seq(len - n + 1, len)]
}

ctb <- ctb %>%
    rowwise() %>%
    mutate(Colors = list(last_n(Colors, Num)))

ctb <- ctb %>%
    select(Phylum, Families, Colors) %>%
    unnest() %>%
    arrange(Phylum, Families)

colors.families <-  ctb$Colors
names(colors.families) <- ctb$Families
# RColorBrewer::display.brewer.all()
# RColorBrewer::brewer.pal(5, name = "Blues")
```

```{r, eval = FALSE, include = FALSE}
# Composition with new colors:
p.nat_comp <- ggplot(nat0,
    aes(x = factor(Individual, as.character(c(1:8, "M"))),
        y = Abundance, 
        fill = factor(Family, names(colors.families)))) +
    geom_bar(stat = "identity") +
    scale_y_continuous(limits=c(0,1), breaks = c(0, 0.5, 1), 
        labels = c("0", "0.5", "1")) +
    geom_rangeframe(data = tibble(y = c(0, 1), Protocol = "H"), 
        aes(y = y), sides = "l", inherit.aes = FALSE) +
    scale_fill_manual(values = colors.families) +
    facet_grid(. ~ Protocol) +
    labs(x = "Specimen",
        title = paste0("Composition of native bacterial taxa")
        ) +
    base_theme +
    theme(legend.position = "right") +
    labs(fill = "Family", y = "Proportion")
p.nat_comp
```

We also want the mock and contaminant fractions in the final figure
```{r}
comp <- bind_rows(
    nat0,
    tb.source %>% filter(Source != "Native") %>% rename(Family = Source)
)
comp %>%
    filter(Sample == "W1") %>%
    select(Family, Abundance) %>%
    arrange(desc(Abundance))
comp %>%
    group_by(Sample) %>%
    summarize(sum= sum(Abundance)) %>%
    {min(.$sum)}
colors.source <- c("#A9A9A9", "#D3D3D3")
names(colors.source) <- c("Contaminant", "Mock")
colors.comp <- c(colors.source, colors.families)
```

```{r, fig.height = 5}
p.comp <- ggplot(comp,
    aes(x = factor(Individual, as.character(c(1:8, "M"))),
        y = Abundance, 
        fill = factor(Family, names(colors.comp)))) +
    geom_bar(stat = "identity") +
    scale_y_continuous(limits=c(0,1), breaks = c(0, 0.5, 1), 
        labels = c("0", "0.5", "1")) +
    geom_rangeframe(data = tibble(y = c(0, 1), Protocol = "H"), 
        aes(y = y), sides = "l", inherit.aes = FALSE) +
    scale_fill_manual(values = colors.comp) +
    facet_grid(. ~ Protocol) +
    labs(x = "Specimen", y = "Proportion",
        title = paste0("Composition of bacterial taxa")
        ) +
    base_theme +
    theme(legend.position = "right", legend.key.size = unit(0.15, "in")) +
    labs(fill = "Source or\nFamily (Native)")
p.comp
```

# Relative abundances and bias in the mock spike-in

Next, we want to visualize relative abundances vs. actual composition (bias) of
the mock. It is useful to plot the abundance of individual taxa as their ratio
to the geometric mean of all taxa in the sample. However, this requires that
all taxa have a positive measured abundance, and there are a few cases where
the rarest mock taxon Blautia hansenii is not detected:
```{r}
st0 %>%
    filter(Taxon %in% mock$Taxon, Abundance == 0)
```
In fact, several reads were assigned to Blautia hansenii in each of these
samples, but not enough to pass Metaphlan2's minimum abundance filter
(Metaphlan's `clade_profiles` output, not shown).  A common way to remove zeros
is add a pseudocount of 1 to all read counts. But Metaphlan provides estimated
proportions rather than read counts, and as noted the detection threshold for
Metaphlan to decide a taxon is present is greater than 1 read anyways.
Instead, we estimate the minimum abundance threshold for each sample from the
proportion of the rarest detected taxon, and then use this as our pseudocount
to add to all taxa of the sample.
```{r}
sample_min <- st0 %>%
    group_by(Sample) %>%
    filter(Abundance > 0) %>%
    summarize(Min = min(Abundance))
sample_min$Min
```
We will further restrict to just the mock taxa,
```{r}
stm <- st0 %>%
    filter(Taxon %in% mock$Taxon) %>%
    left_join(sample_min, by = "Sample") %>%
    mutate_by(Sample, Abundance = close_elts(Abundance + Min)) %>%
    select(-Min)
stm %>%
    filter(Taxon == "Blautia_hansenii") %>%
    arrange(Abundance)
```
Add the actual abundances, (Taking the FACS measurement to give the actual
relative abundances), and rename Abundance to Observed.
```{r}
stm <- stm %>%
    left_join(mock, by = "Taxon") %>%
    mutate_by(Sample, Actual = close_elts(FACS)) %>%
    rename(Observed = Abundance)
```

From now on, we will work with the geometrically centered relative abundances.
```{r}
stm <- stm %>%
    mutate_by(Sample, 
        Actual = center_elts(Actual),
        Observed = center_elts(Observed)
    )
```

Plot the observed and actual relative abundances.
```{r, fig.width = 5, fig.asp = 3/4}
p.step <- ggplot(stm,
    aes(x = factor(Taxon, mock$Taxon), y = Observed,
        color = Protocol, shape = (Individual == "M"))) + 
    geom_quasirandom() +
    geom_rangeframe(color = "black", sides = "l") +
    scale_color_manual(values = colors.protocol) + 
    # scale_y_log10(breaks = c(0.003, 1e-2, 1e-1, 1, 10, 30),
    #     labels = c("3e-3", "1e-2", "0.1", "1", "10", "30")) +
    scale_y_log10(labels = log_formatter) +
    scale_x_discrete(labels = tax_labeller) + 
    costea.shape_scale + 
    scale_color_manual(values = colors.protocol) + 
    coord_cartesian(clip = "off") +
    labs(title = "Relative abundances of mock taxa",
        y = "Abundance relative to geometric mean",
    shape = "Specimen type") +
    base_theme + 
    tax_theme + 
    theme(legend.position = "right")
# Add the actual abundances (as a stair plot like in Costea2017's Figure 6)
# underneath the data points
tb.step <- stm %>%
    filter(Sample == "H1") %>%
    select(Taxon, Actual) %>%
    mutate(x1 = rank(desc(Actual)) - 0.5, x2 = x1 + 0.5,
        x3 = ifelse(x2 == 10, 10.5, NA)) %>%
    gather("loc", "x", x1, x2, x3) %>%
    filter(!is.na(x)) %>%
    arrange(x)
p.step$layers <- c(
    geom_step(data = tb.step, aes(x = x, y = Actual), 
        size = 0.3, color = "black", inherit.aes = FALSE),
    p.step$layers)
p.step <- p.step +
    annotate("text", x = 10.6, y = 0.025, label = "Actual",
        color = "black", hjust = 0)
p.step
```

## Main text figure

```{r, fig.width = 7, fig.height = 3.5}
plot_grid(p.comp, p.step, labels = c("A", "B"))
```

```{r, include = FALSE, eval = SAVE_FIGURES}
# ggsave(file.path(here::here(), "figures", "costea2017_2panel.svg"),
#     width = 7, height = 3.5, units = "in", scale = 4/3)
ggsave(file.path(here::here(), "figures", "costea2017_2panel.png"),
    width = 7, height = 3.5, units = "in", dpi = 300, scale = 4/3)
ggsave(file.path(here::here(), "figures", "costea2017_2panel.pdf"),
    width = 7, height = 3.5, units = "in", scale = 4/3)
```

# Estimate bias

Estimate the bias separately within each protocol using the "center" method of
the `estimate_bias` function,
```{r}
bias <- stm %>%
    group_by(Protocol) %>%
    nest %>%
    mutate(Bias = map(data, estimate_bias, method = "center")) %>%
    unnest(Bias) %>%
    rename(Bias = Bias_est)
bias %>%
    spread(Protocol, Bias)
```
The estimated bias vectors are geometrically centered by dividing by the
geometric mean of taxa,
```{r}
bias %>%
    group_by(Protocol) %>%
    summarize(gm_mean(Bias))
```
so that the elements (relative efficiencies) can be interpreted as the
detection efficiency relative to the average taxon for that protocol.

To confirm that our estimate is good, we can use it to calibrate the relative
abundances in the previous figure:
```{r, fig.width = 5, fig.asp = 3/4}
p.step.cal <- p.step
p.step.cal$data <- p.step.cal$data %>%
    left_join(bias, by = c("Protocol", "Taxon")) %>%
    mutate_by(Sample, Observed = center_elts(Observed / Bias))
p.step.cal
```

Check that the bias predictions predict the taxon ratios:
```{r}
ratios <- crossing(Sample = setdiff(sam$Sample, c("QA", "QB")), 
    Taxon.x = mock$Taxon, Taxon.y = mock$Taxon) %>%
    left_join(stm, by = c("Sample", "Taxon.x" = "Taxon")) %>%
    left_join(stm %>% select(Sample, Taxon, Observed, Actual), 
        by = c("Sample", "Taxon.y" = "Taxon")) %>%
    mutate(Pair = paste(Taxon.x, Taxon.y, sep = ":"))
ratios <- ratios %>%
    left_join(bias, by = c("Protocol", "Taxon.x" = "Taxon")) %>%
    left_join(bias, by = c("Protocol", "Taxon.y" = "Taxon"))
ratios <- ratios %>%
    mutate(
        Observed = Observed.x / Observed.y,
        Actual = Actual.x / Actual.y,
        Error = Observed / Actual,
        Bias = Bias.x / Bias.y,
        Predicted = Actual * Bias,
    )
ratios.pred <- ratios %>%
    select(Taxon.x, Taxon.y, Pair, Protocol, Actual, Bias, Predicted) %>%
    distinct
```
```{r, fig.width = 10, fig.height = 7}
ggplot(ratios, aes(Pair, Error, 
        label = Individual, color = Protocol)) +
    geom_point(data = ratios.pred,
        aes(Pair, Bias), inherit.aes = FALSE,
        shape = 3, size = 4, color = "black") +
    geom_text(position = position_jitter(width=0.25, height=0)) +
    geom_hline(yintercept = 1) +
    facet_grid(Protocol ~ .) +
    geom_rangeframe(sides = "l", color= "black") + 
    scale_color_manual(values = colors.protocol) + 
    scale_y_log10() +
    scale_x_discrete(labels = tax_labeller) + 
    base_theme +
    tax_theme
```

Distribution of efficiencies for the three protocols:
```{r, fig.height = 2, fig.width = 3}
ggplot(bias, aes(Bias, fill = Protocol)) +
    geom_dotplot() +
    scale_x_log10() +
    facet_grid(Protocol ~ .) +
    scale_fill_manual(values = colors.protocol) +
    scale_y_continuous(breaks = NULL) +
    geom_rangeframe() +
    base_theme
```

Make a table for the main text using kable and kableExtra:
```{r}
bias_summary <- bias %>%
    group_by(Protocol) %>%
    summarize(
        `Gm. range` = max(Bias) / min(Bias),
        `Gm. std. dev.` = exp(sd(log(Bias)))
    ) %>%
    gather("Statistic", "Value", -Protocol) %>%
    spread(Protocol, Value)
bias_summary
lvls <- c(mock$Taxon, "Gm. range", "Gm. std. dev.")
bias_tab <- bind_rows(
    bias %>% spread(Protocol, Bias),
    bias_summary %>% rename(Taxon = Statistic)
    ) %>%
    mutate(Taxon = factor(Taxon, lvls)) %>%
    arrange(Taxon)
bias_tab
```

Table for the main text
```{r}
tex <- bias_tab %>%
    # mutate(Taxon = str_replace(Taxon, "_", " ")) %>%
    mutate(
        Taxon = kableExtra::cell_spec(
            str_replace(Taxon, "_", " "), 
            "latex", 
            italic = Taxon %in% mock$Taxon)
        ) %>%
    mutate_at(vars(-Taxon), signif, 2) %>%
    mutate_at(vars(-Taxon), as.character) %>%
    knitr::kable(format="latex", booktabs = TRUE, linesep = "",
        escape = FALSE, align = c("l", "r", "r", "r")) %>%
    # knitr::kable(digits=1, format="latex", booktabs = TRUE, linesep = "",
    #     escape = FALSE) %>%
    # kableExtra::kable_styling(latex_options = "striped")
    kableExtra::row_spec(10, extra_latex_after = "\\midrule")
tex
```
<!--
TODO: Edit the above to show 2 sig figures instead of 1 decimal place (esp. to
deal with L plantarum in protocol H).
TODO: consider adding phylum and/or gram type
-->

# Calibration

Choose Individuals (specimens) to use for estimating bias.
```{r}
set.seed(20190201)
individuals <- sam$Individual %>% unique
estimation_set <- base::sample(individuals, 3, replace = FALSE)
stm0 <- stm %>%
    mutate(Set = ifelse(Individual %in% estimation_set, "Est", "Eval"))
sam <- sam %>%
    mutate(Set = ifelse(Individual %in% estimation_set, "Est", "Eval"))
```
Estimate bias from the estimation samples:
```{r}
bias_est <- stm0 %>%
    filter(Set == "Est") %>%
    group_by(Protocol) %>%
    nest %>%
    mutate(Bias = map(data, estimate_bias, method = "center")) %>%
    unnest(Bias) %>%
    rename(Bias = Bias_est)
```
Compare the bias estimated from the estimation set and that from all specimens:
```{r}
bind_rows(list(All = bias, Est = bias_est), .id = "Specimens") %>%
    ggplot(aes(Taxon, Bias, color = Protocol, shape = Specimens)) + 
    geom_quasirandom() +
    scale_y_log10() +
    scale_shape_manual(breaks = c("All", "Est"), values = c(16, 1)) +
    scale_color_manual(values = colors.protocol) + 
    coord_flip()
```

Bias is consistent with the estimate using all specimens, though this would
vary for the higher-variance protocols H and Q for some estimation sets.

Also estimate the differential bias to protocol W. First, add a column with the
"Reference" measurement; then repeat bias estimation with the reference
measurement instead of the Actual abundance.
```{r}
stmW <- stm0 %>%
    select(Taxon, Protocol, Individual, Set, Observed) %>%
    spread(Protocol, Observed) %>%
    mutate(Actual = W) %>%
    gather("Protocol", "Observed", H, Q, W)
biasW_est <- stmW %>%
    filter(Set == "Est") %>%
    group_by(Protocol) %>%
    nest %>%
    mutate(Bias = map(data, estimate_bias, method = "center")) %>%
    unnest(Bias) %>%
    rename(BiasW = Bias_est)
biasW_est %>%
    spread(Protocol, BiasW)
```

Calibrate all samples
```{r}
stm0 <- stm0 %>%
    left_join(bias_est, by = c("Protocol", "Taxon")) %>%
    left_join(biasW_est, by = c("Protocol", "Taxon")) %>%
    mutate_by(Sample, 
        Calibrated_to_Actual = close_elts(Observed / Bias),
        Calibrated_to_W = close_elts(Observed / BiasW)
    )
```

## Sample ordination: Compositional PCA

Prep for the PCA: we first need to Clr transform each relative abundance
vector.
```{r}
stm1 <- stm0 %>%
    gather("Type", "Abundance", 
        Observed, Actual, Calibrated_to_Actual, Calibrated_to_W) %>%
    select(Taxon:Individual, Set:Abundance) %>%
    group_by(Sample, Type) %>%
    mutate(Clr = clr(Abundance)) %>%
    ungroup
```
Get a matrix with samples as rows to use for the PCA.
```{r}
temp <- stm1 %>%
    unite("Sample_Type", Sample, Type, sep = ":") %>%
    select(Sample_Type, Taxon, Clr) %>%
    spread(Taxon, Clr)
mat <- temp %>%
    select(-Sample_Type) %>%
    as("matrix")
rownames(mat) <- temp$Sample_Type
corner(mat)
```
Note that each sample (protocol + individual) appears four times, as Observed,
Actual, Calibrated to Actual, and Calibrated to W, and currently I'm just doing
a simple PCA on all of these, since our goal is mainly to illustrate the effect
of calibration rather than make quantitative claims about variation. Next we
run the PCA and make a tibble for plotting,
```{r}
pcx <- prcomp(mat)
tb <- as_tibble(pcx$x[,c(1,2)], rownames='Sample_Type') %>%
    separate(Sample_Type, c("Sample", "Type"), sep = ":") %>%
    left_join(sam, by = "Sample")
tb <- tb %>%
    mutate(PC1 = -PC1, PC2 = PC2)
```
Nearly 90% of the variance is covered by the first two PCs, with about 3.4x as
much variation on PC1 than PC2:
```{r}
fracvar <- pcx$sdev^2/sum(pcx$sdev^2)
percvar <- round(100 * fracvar, 1)
strivar <- paste0("PC", seq(fracvar), " [", percvar, "%]")
strivar
```

PCA figure that will form the right half of the main text figure:
```{r}
facet_labeller <- function (tb) {
    tb %>% 
        mutate_all(str_replace_all, "_", " ") %>%
        mutate_all(str_replace, " ", "\n")
}
type.lvls <- c("Observed", "Calibrated_to_Actual", "Calibrated_to_W")
actual <- tb %>%
    filter(Type == "Actual", Sample == "H1") %>%
    select(PC1, PC2) %>%
    expand(Type = factor(type.lvls, type.lvls), PC1, PC2)
tb0 <- tb %>%
    filter(Type != "Actual") %>%
    mutate(Type = factor(Type, type.lvls))
tb.breaks <- 
    tribble(
        ~PC1, ~PC2,
        min(tb0$PC1), min(tb0$PC2),
        actual$PC1[1], actual$PC2[1],
        max(tb0$PC1), max(tb0$PC2),
        )
p.pca <- ggplot(tb0, aes(PC1, PC2, color = Protocol, shape = Set)) +
    geom_point(data = actual, aes(PC1, PC2),
        shape = 3, size = 4, color = "black") +
    geom_point() +
    facet_grid(Type ~ ., labeller = facet_labeller) +
    geom_rangeframe(data = tb.breaks, aes(PC1, PC2), 
        color = "black", inherit.aes = FALSE) + 
    scale_x_continuous(breaks = tb.breaks$PC1, labels = NULL) +
    scale_y_continuous(breaks = tb.breaks$PC2, labels = NULL) +
    scale_shape_manual(breaks = c("Est", "Eval"), values = c(1, 16)) +
    scale_color_manual(values = colors.protocol) + 
    base_theme +
    theme(strip.text.y = element_text(angle=0, size = 9)) +
    # theme(axis.line = element_line()) +
    labs(title = "Sample ordination")
# Label the protocols in the "Observed" plot
labtb <- tribble(
    ~x,     ~y,   ~Protocol,
    -4,    1.5,   "H",
    0,     2,     "Q",
    -1.3,  -0.5,  "W",
    2.3,     0.9,   "Actual"
    ) %>%
    mutate(Type = factor("Observed", type.lvls))
p.pca <- p.pca +
    geom_text(data = labtb,
        aes(x, y, label = Protocol, color = Protocol), 
        size = 3, inherit.aes = FALSE, show_guide = FALSE)
# p.pca
```

## Relative abundances before and after calibration

Figure showing the relative abundances, which will form the left half of the
main text figure:
```{r}
stm1 <- stm1 %>% 
    group_by(Sample, Type) %>%
    mutate(Abundance = center_elts(Abundance)) %>%
    ungroup
tbrf <- tibble(Abundance = range(stm1$Abundance))
p.step.cal <- stm1 %>% 
    filter(Type %in% type.lvls) %>%
    mutate(Taxon = factor(Taxon, mock$Taxon),
        Type = factor(Type, type.lvls)) %>%
    ggplot(aes(x = Taxon, y = Abundance, color = Protocol, shape = Set)) + 
    geom_quasirandom() +
    geom_rangeframe(data = tbrf, aes(y = Abundance), 
        color = "black", sides = "l", inherit.aes = FALSE) +
    scale_color_manual(values = colors.protocol) + 
    scale_y_log10(labels = log_formatter) +
    scale_x_discrete(labels = tax_labeller) + 
    coord_cartesian(clip = "off") +
    labs(title = "Taxon relative abundances",
        y = "Abundance relative to geometric mean") + 
    facet_grid(Type ~ .) +
    scale_shape_manual(breaks = c("Est", "Eval"), values = c(1, 16)) +
    scale_color_manual(values = colors.protocol) + 
    base_theme + 
    tax_theme +
    theme(strip.text.y = element_blank())
# Add the actual abundances (as a stair plot like in Costea2017's Figure 6)
# underneath the data points
tb.step.cal <- tb.step %>%
    expand(Type = factor(type.lvls, type.lvls), nesting(x, Actual))
p.step.cal$layers <- c(
    geom_step(data = tb.step.cal, aes(x = x, y = Actual), 
        size = 0.3, color = "black", inherit.aes = FALSE),
    p.step.cal$layers)
# p.step.cal
```

## Main text figure

```{r, fig.width = 6.5, fig.height = 6.5}
plot_grid(p.step.cal, 
    p.pca + 
        theme(legend.position = "bottom", legend.box = "vertical",
            legend.margin = margin(),
            legend.spacing = unit(0, "in"),
        )
)
```

```{r, include = FALSE, eval = SAVE_FIGURES}
ggsave(file.path(fig_path, "costea2017_calibration.png"),
    width = 6.5, height = 6.5, units = "in")
ggsave(file.path(fig_path, "costea2017_calibration.pdf"),
    width = 6.5, height = 6.5, units = "in")
```

# Session info

```{r}
sessionInfo()
```
